The classical mutual information in mean-field spin glass models

### 2015-12-10 sk_model_v20.tex
rev-10/11/2015

\documentclass[twocolumn,superscriptaddress,prb,10pt]{revtex4-1}
%\usepackage{verbatim}
\usepackage{amsmath,amssymb}
\usepackage{graphicx}
\usepackage{color}
\usepackage[colorlinks,bookmarks=false,citecolor=blue,linkcolor=red,urlcolor=blue]{hyperref}
\usepackage{times}

@@

We start discussing the thermodynamic phase diagram of the S-K model on the $n$-sheets 
booklet, as a function of temperature, and the aspect ratio $\omega$ (cf. Eq.~\eqref{a-ratio}). 
For any fixed $\omega$ the S-K model exhibits a low-temperature glassy phase, which is divided 
by the standard paramagnetic one at high temperatures by a phase transition. 
%%%
The critical 
inverse temperature $\beta_c$ exhibits a non trivial decreasing behavior as a function of 
$\omega$ that we are able to determine analytically. Moreover, one has $\beta_c=1$ 
and $\beta_c=1/2$ for $\omega=0$ and $\omega=1$, respectively. 
%%%
In the high-temperature region the permutation symmetry among both the replicas  
and the physical sheets  is preserved. This allows us to provide an exact 
analytic expression for the free energy of the model and several derived quantities, such as 
the internal energy. We compare our results with Monte Carlo simulations, finding perfect 
agreement. Oppositely, in the low-temperature phase the replica symmetry is broken. For 
instance, we numerically observe that the replica-symmetric (RS) result for the internal 
energy is systematically lower than the Monte Carlo data, as in the standard S-K 
model~\cite{sherrington-1978-prl,sherrington-1978}. This discrepancy becomes larger upon 
lowering the temperature.

@@

Inspired by the Parisi scheme~\cite{parisi-1979}, we devise a systematic way of breaking 
the replica symmetry in successive steps. In our scheme we break only the symmetry among 
the fictitious replicas, preserving that among the physical ones. Although this appears 
natural, we are not able to provide a rigorous proof that this is the correct symmetry 
breaking pattern. As a consequence, our scheme should be regarded as an approximation, 
and not as an exact solution. Moreover, we restrict ourselves to the one-level replica 
symmetry breaking ($1$-RSB). Surprisingly, the $1$-RSB result for the internal energy 
is in excellent agreement with the Monte Carlo data for  $\beta\lesssim 3$, whereas the 
RS approximation fails already at $\beta\approx 1$. This suggests that the $1$-RSB ansatz 
captures correctly some aspects of the replica symmetry breaking.

Clear signatures of the replica symmetry breaking are observed in the behavior of $[S_2]$. 
First, since $[S_2]$ exhibits the volume law behavior $[S_2]\propto N$, 
we consider its density $[S_2]/N$. For finite-size systems, and for any $\omega$, $[S_2]/N$ 
exhibits a maximum at infinite temperature, and it is a decreasing function of the temperature, 
as expected. Finite-size corrections are negligible at high temperatures, whereas they increase 
upon lowering the temperature. In the paramagnetic phase we are able to determine the functional 
form of $[S_2]/N$ in the thermodynamic limit, using the replica-symmetric approximation. This 
perfectly matches the Monte Carlo data. At low temperatures deviations from the RS result are 
present, reflecting the replica symmetry breaking. Remarkably, the one-step replica symmetry 
breaking ($1$-RSB) result $[S_2^{1\textrm{-}RSB}]/N$ is in full agreement with the Monte Carlo 
data for $\beta\lesssim 3$, consistent with what is observed for the internal energy.

Finally, we consider the R\'enyi mutual information $[{\mathcal I}_2]$. This obeys a volume 
law for any $\beta$ and $\omega$, in contrast with local spin models, where an area law is 
observed~\cite{wolf-2008}. The corresponding density $[{\mathcal I}_2]/N$ vanishes in 
both the infinite-temperature and the zero-temperature limits. Surprisingly, $[{\mathcal I}_2]/N$ 
does not exhibit any crossing for different system sizes at the paramagnetic-glassy transition, 
in striking contrast with local spin models~\cite{jaconis-2013}. For any $\omega$, $[{\mathcal I}_2]/N$ 
exhibits a maximum for $\beta\approx 1$. The position of this maximum is not simply 
related to the paramagnetic-glassy transition. At high temperature $[{\mathcal I}_2]/N$ 
is described analytically by the RS result $[{\mathcal I}_2]^{RS}/N$. Deviations from the RS result, 
if present, are $\lesssim 10^{-3}$. On the other hand, at low temperatures 
one has to include the effects of the replica symmetry breaking. Similar to $[S_2]$, the $1$-RSB 
approximation $[{\mathcal I}^{1\textrm{-}RSB}_2]/N$ is in good agreement with the Monte Carlo 
data for $\beta\lesssim 3$.

@@

The partition function $Z(\omega,n,\beta,\{J\})$ of the S-K model on the booklet 
at inverse temperature $\beta\equiv 1/T$, and for fixed disorder realization 
$\{J_{ij}\}$, reads 
%
\begin{multline}
Z(\omega,n,\beta,\{J\})\equiv\textrm{Tr}'\exp(-\beta{\mathcal H})=\\
\textrm{Tr}'\exp\Big\{\beta\sum\limits_{r}\Big(
\sum\limits_{i<j}S^{(r)}_i 
S^{(r)}_j-h\sum\limits_{i=1}^NS^{(r)}_i\Big)\Big\},
\label{book-pf}
\end{multline}
%
where $\textrm{Tr}'\equiv \sum_{\{S_i\}}$ denotes the sum over all possible 
spin configurations. The prime in $\textrm{Tr}'$ stresses that only spin configurations 
satisfying the constraint in Eq.~\eqref{book-constraint} are considered. In the two limits 
$\omega=0$ and $\omega=1$ one recovers the standard S-K model. In particular, for $\omega=0$, 
i.e., $n$ disconnected sheets, one has $Z(0,n,\beta,\{J\})=Z(\beta,\{J\})^n$, with 
$Z(\beta,\{J\})$ the partition function of the S-K model on the plane (i.e., the original 
S-K model). On the other hand, 
for $\omega=1$ it is $Z(1,n,\beta,\{J\})=Z(n\beta,\{J\})$, i.e., the partition function 
of the S-K model at inverse temperature $n\beta$. The quenched averaged free energy 
$[F(\omega,n,\beta)]$, is defined as 
%
\begin{equation}
\label{free-energy}
[F(\omega,n,\beta)]\equiv -\frac{1}{\beta}\int {\mathcal D}\{J\}
\log Z(\omega,n,\beta,\{J\}), 
\end{equation}
%
where $\int{\mathcal D}\{J\}\equiv\prod\nolimits_{i<j}\int_{-\infty}^{
+\infty}dJ_{ij}P(J_{ij})$.

@@

We investigated the \emph{classical} R\'enyi entropy $S_n$ and the mutual information 
${\mathcal I}_n$ in the Sherrington-Kirkpatrick (S-K) model, which is the paradigm model 
of mean-field spin glasses. We focused on the quenched averages $[S_n]$ and $[{\mathcal I}_n]$. 
Specifically, here $[S_n]$ and $[{\mathcal I}_n]$ were obtained from suitable combinations 
of the partition functions of the S-K model on the $n$-sheets booklet (cf. Fig.~\ref{cartoon}). 
This is constructed by ``gluing'' together $n$ independent replicas (``sheets'') of the model. 
On each replica the spins are divided into two groups $A$ and $B$, containing $N_A$ and $N_B$ 
spins respectively. The spins in part $A$ of the different sheets are identified. Due to the 
mean-field nature of the model, physical quantities depend on the bipartition only through the 
aspect ratio $\omega\equiv N_A/N$.

@@

We first investigated the phase diagram of the S-K model on the $2$-sheets booklet as a 
function of $\omega$, and the inverse temperature $\beta$. We considered the situation 
without external magnetic field ($h=0$ in Eq.~\eqref{SK-intro}). For any $\omega$, in the 
thermodynamic limit, the model exhibits a glassy phase for low enough temperatures and 
a standard paramagnetic one at high temperature. The two phases are divided 
by a phase transition, as in the S-K model on the plane. The critical temperature exhibits 
a non-trivial dependence on $\omega$ that we determined analytically. We fully characterized 
the high-temperature phase of the model using the replica-symmetric (RS) ansatz. Specifically, 
we provided the exact analytical expression for the disorder-averaged free energy 
$[F_{para}(\omega,n,\beta)]$  (cf. Eq.~\eqref{ht-logZ}) in the thermodynamic limit. In the 
glassy phase the replica symmetry has to be broken, similar to the standard S-K model. To take 
into account the effects of the replica symmetry breaking we used a generalization of the 
Parisi ansatz~\cite{parisi-1980}, i.e., breaking the replica symmetry in successive steps. 
We restricted ourselves to the one-step replica symmetry breaking ($1$-RSB), providing the 
exact expression for the disorder-averaged free energy $F_{1\textrm{-}RSB}(\omega,n,\beta)$ 
(cf. Eq.~\eqref{RSB-1-logZ}).

We compared our analytical results with Monte Carlo simulations, focusing on the internal 
energy per spin $U(\omega,n,\beta)/N$. In the high-temperature region, where the replica 
symmetry is not broken, the Monte Carlo data are in perfect agreement with the replica-symmetric 
result $U_{RS}/N$, for any $\omega$. In the glassy phase at low temperature, significant 
deviations from $U_{RS}$ can be observed already at $\beta\approx 1$. On the other hand, 
the one-step replica-symmetry-broken approximation $U_{1\textrm{-}RSB}/N$ accurately matches 
the Monte Carlo data for $\beta\lesssim 3$.

Finally, we discussed the behavior of $[S_2]$ and $[{\mathcal I}_2]$. Due to the mean-field 
nature of the S-K model one has the volume laws $[S_2]\propto N$ and $[{\mathcal I}_2]\propto N$, 
for any $\beta$. We considered the densities $[S_2]/N$ and $[{\mathcal I}_2]/N$. Using the 
replica-symmetric (RS) and the one-step replica symmetry breaking ($1$-RSB) approximations we 
provided analytical results for both $[S_2]/N$ and $[{\mathcal I}_2]/N$. At high temperature the 
RS result fully agrees with the Monte Carlo simulations, whereas deviations occur in the 
glassy phase, signalling the replica symmetry breaking. Similar to the internal energy, 
the $1$-RSB approximation fully matches the Monte Carlo data for $\beta\approx 3$.
Interestingly, at the critical point $[{\mathcal I}_2]/N$ does not show any crossing for 
different system sizes, in contrast with local models~\cite{jaconis-2013}.

@@

>>>>

\documentclass[twocolumn,superscriptaddress,prb,10pt]{revtex4-1}
\usepackage{verbatim}
\usepackage{amsmath,amssymb}
\usepackage{graphicx}
\usepackage{color}
\usepackage[colorlinks,bookmarks=false,citecolor=blue,linkcolor=red,urlcolor=blue]{hyperref}
\usepackage{times}

@@

@@

@@

The partition function $Z(\omega,n,\beta,\{J\})$ of the S-K model on the booklet 
at inverse temperature $\beta\equiv 1/T$, and for fixed disorder realization 
$\{J_{ij}\}$, reads 
%
\begin{multline}
Z(\omega,n,\beta,\{J\})\equiv\textrm{Tr}'\exp(-\beta{\mathcal H})=\\
\textrm{Tr}'\exp\Big\{\beta\sum\limits_{r}\Big(
\sum\limits_{i<j}J_{ij}S^{(r)}_i 
S^{(r)}_j-h\sum\limits_{i=1}^NS^{(r)}_i\Big)\Big\},
\label{book-pf}
\end{multline}
%
where $\textrm{Tr}'\equiv \sum_{\{S_i\}}$ denotes the sum over all possible 
spin configurations. The prime in $\textrm{Tr}'$ stresses that only spin configurations 
satisfying the constraint in Eq.~\eqref{book-constraint} are considered. In the two limits 
$\omega=0$ and $\omega=1$ one recovers the standard S-K model. In particular, for $\omega=0$, 
i.e., $n$ disconnected sheets, one has $Z(0,n,\beta,\{J\})=Z(\beta,\{J\})^n$, with 
$Z(\beta,\{J\})$ the partition function of the S-K model on the plane (i.e., the original 
S-K model). On the other hand, 
for $\omega=1$ it is $Z(1,n,\beta,\{J\})=Z(n\beta,\{J\})$, i.e., the partition function 
of the S-K model at inverse temperature $n\beta$. The quenched averaged free energy 
$[F(\omega,n,\beta)]$, is defined as 
%
\begin{equation}
\label{free-energy}
[F(\omega,n,\beta)]\equiv -\frac{1}{\beta}\int {\mathcal D}\{J\}
\log Z(\omega,n,\beta,\{J\}), 
\end{equation}
%
where $\int{\mathcal D}\{J\}\equiv\prod\nolimits_{i<j}\int_{-\infty}^{
+\infty}dJ_{ij}P(J_{ij})$.

@@

We investigated the \emph{classical} R\'enyi entropy $S_n$ and the mutual information 
${\mathcal I}_n$ in the Sherrington-Kirkpatrick (S-K) model, which is the paradigm model 
of mean-field spin glasses. We focused on the quenched averages $[S_n]$ and $[{\mathcal I}_n]$. 
Specifically, here $[S_n]$ and $[{\mathcal I}_n]$ are obtained from suitable combinations 
of the partition functions of the S-K model on the $n$-sheets booklet (cf. Fig.~\ref{cartoon}). 
This is constructed by ``gluing'' together $n$ independent replicas (``sheets'') of the model. 
On each replica the spins are divided into two groups $A$ and $B$, containing $N_A$ and $N_B$ 
spins respectively. The spins in part $A$ of the different sheets are identified. Due to the 
mean-field nature of the model, physical quantities depend on the bipartition only through the 
aspect ratio $\omega\equiv N_A/N$.

@@

We first discussed the thermodynamic phase diagram of the S-K model on the $n$-sheets 
booklet, as a function of temperature, and the aspect ratio $\omega$ (cf. Eq.~\eqref{a-ratio}). 
For any fixed $\omega$ the S-K model exhibits a low-temperature glassy phase, which is divided 
by the standard paramagnetic one at high temperatures by a phase transition. 
%%%
The critical inverse temperature $\beta_c$ exhibits a non trivial decreasing behavior as a 
function of $\omega$. Moreover, one has $\beta_c=1$ and $\beta_c=1/2$ for $\omega=0$ and 
$\omega=1$, respectively. 
%%%
In the high-temperature region the permutation symmetry among both the replicas  and the physical 
sheets  is preserved. This allowed us to provide an exact analytic expression for the free 
energy of the model and several derived quantities, such as the internal energy. 
We compared our results with Monte Carlo simulations, finding perfect agreement. Oppositely, 
in the low-temperature phase the replica symmetry is broken. For instance, we numerically 
observed that the replica-symmetric (RS) result for the internal energy is systematically 
lower than the Monte Carlo data, as in the standard S-K model~\cite{sherrington-1978-prl,
sherrington-1978}. This discrepancy becomes larger upon lowering the temperature.

Inspired by the Parisi scheme~\cite{parisi-1979}, we devised a systematic way of breaking 
the replica symmetry in successive steps. Our scheme breaks only the symmetry among 
the fictitious replicas, preserving that among the physical ones. Although this appears 
natural, we were not able to provide a rigorous proof that this is the correct symmetry 
breaking pattern. As a consequence, our scheme should be regarded as an approximation, 
and not as an exact solution. Moreover, we restricted ourselves to the one-level replica 
symmetry breaking ($1$-RSB). Surprisingly, the $1$-RSB result for the internal energy 
were in excellent agreement with the Monte Carlo data for  $\beta\lesssim 3$, whereas the 
RS approximation failed already at $\beta\approx 1$. This suggests that the $1$-RSB ansatz 
captures correctly some aspects of the replica symmetry breaking.

Clear signatures of the replica symmetry breaking can be observed in the behavior of $[S_2]$. 
First, since $[S_2]$ exhibits the volume law behavior $[S_2]\propto N$, we considered its 
density $[S_2]/N$. For finite-size systems, and for any $\omega$, $[S_2]/N$ 
exhibits a maximum at infinite temperature, and it is a decreasing function of the temperature, 
as expected. Finite-size corrections are negligible at high temperatures, whereas they increase 
upon lowering the temperature. In the paramagnetic phase we are able to determine the functional 
form of $[S_2]/N$ in the thermodynamic limit, using the replica-symmetric approximation. This 
perfectly matched the Monte Carlo data. At low temperatures deviations from the RS result are 
present, reflecting the replica symmetry breaking. Remarkably, the one-step replica symmetry 
breaking ($1$-RSB) result $[S_2^{1\textrm{-}RSB}]/N$ fully described the Monte Carlo 
data for $\beta\lesssim 3$, consistent with what was observed for the internal energy.

Finally, we considered the R\'enyi mutual information $[{\mathcal I}_2]$. This obeys a volume 
law for any $\beta$ and $\omega$, in contrast with local spin models, where an area law is 
observed~\cite{wolf-2008}. The corresponding density $[{\mathcal I}_2]/N$ vanishes in 
both the infinite-temperature and the zero-temperature limits. Surprisingly, $[{\mathcal I}_2]/N$ 
does not exhibit any crossing for different system sizes at the paramagnetic-glassy transition, 
in striking contrast with local spin models~\cite{jaconis-2013}. For any $\omega$, $[{\mathcal I}_2]/N$ 
exhibits a maximum for $\beta\approx 1$. The position of this maximum is not simply 
related to the paramagnetic-glassy transition. At high temperature $[{\mathcal I}_2]/N$ 
is described analytically by the RS result $[{\mathcal I}_2]^{RS}/N$. Deviations from the RS result, 
if present, are $\lesssim 10^{-3}$. On the other hand, at low temperatures 
one has to include the effects of the replica symmetry breaking. Similar to $[S_2]$, the $1$-RSB 
approximation $[{\mathcal I}^{1\textrm{-}RSB}_2]/N$ is in good agreement with the Monte Carlo 
data for $\beta\lesssim 3$.

@@

%##########################################################################
\section{Mutual information in the infinite-range clean Ising model}
\label{clean-is}

Here we discuss the R\'enyi entropies and the mutual information in the 
infinite-range Ising model withoud disorder defined by the hamiltonian 
%
\begin{equation}
\label{is-clean}
{\mathcal H}=-\frac{J}{N}\sum\limits_{i<j}S_iS_j+h\sum_i S_i.
\end{equation}
%
Here we consider the situation without external magnetic field, i.e., 
$h=0$ and fix $J=1$.

The calculation of the booklet partition function $Z(\omega,n,\beta)$ is 
done using the same techniques as in section~\ref{}. The result reads 
%
\begin{multline}
\label{clean-is-Z}
Z(\omega,n,\beta)=
\exp\Big(-\frac{\beta}{2}n\Big)\Big(2\pi\frac{\beta}{N}\Big)^{-\frac{n}{2}}
\int\prod_{\delta}dz_\delta\exp N\Big\{\\
-\sum_\delta\Big[\frac{z^2_\delta}{2\beta}
-(1-\omega)\log(2\cosh z_\delta)\Big]+\omega\log(2\cosh\sum_\delta z_\delta)
\Big\}.
\end{multline}
%
Using the saddle point approximation one obtains 
%
\begin{multline}
\label{clean-is-F}
\log Z(\omega,n,\beta)\approx
N\Big\{-\sum_\delta\Big[\frac{z^2_\delta}{2\beta}
-(1-\omega)\log(2\cosh z_\delta)\Big]\\
+\omega\log(2\cosh\sum_\delta z_\delta)\Big\}, 
\end{multline}
%
where we neglected subleading contributions $o(N)$. The parameters 
$\{z_\delta\}_{\delta=1}^n$ are solutions of the saddle point equations 
%
\begin{equation}
\label{clean-is-spe}
0=z_\delta-\beta(1-\omega)\tanh z_\delta-\beta\omega\tanh\big(\sum_\delta 
z_\delta\big),\quad \forall\delta.
\end{equation}
%
For $n=2$, Eq.~\eqref{clean-is-spe} become 
%
\begin{align}
0=z_1-\beta(1-\omega)\tanh z_1-\beta\omega\tanh\big(z_1+z_2\big)\\
0=z_2-\beta(1-\omega)\tanh z_2-\beta\omega\tanh\big(z_1+z_2\big). 
\end{align}
%
The system has the solution $z_1=z_2=0$ for $\beta\le\beta_c=1/(1+\omega)$, 
and $z_1=z_2\ne 0$ for $\beta>\beta_c$.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%e
\begin{figure}[t]
\includegraphics*[width=0.9\linewidth]{./draft_figs/saddle_clean}
%\includegraphics*[width=0.9\linewidth]{./RSB0_saddle_ht}
\caption{
 The S-K model on the $2$-sheets booklet in the 
 paramagnetic phase: the solution $q_0$ of the saddle point 
 equation~\eqref{saddle-high-t} plotted as a function of the booklet aspect 
 ratio $\omega$ and several values of the inverse temperature $\beta=1/T$. 
 At $\omega=1$ one has $q_0=1,\forall\beta$. In the zero-temperature limit 
 one has $q_0\to 1,\forall\omega$. The straight line is the infinite 
 temperature result. 
}
\label{saddle_clean}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%e
\begin{figure}[t]
\includegraphics*[width=0.9\linewidth]{./draft_figs/renyi_clean}
%\includegraphics*[width=0.9\linewidth]{./RSB0_saddle_ht}
\caption{
 The S-K model on the $2$-sheets booklet in the 
 paramagnetic phase: the solution $q_0$ of the saddle point 
 equation~\eqref{saddle-high-t} plotted as a function of the booklet aspect 
 ratio $\omega$ and several values of the inverse temperature $\beta=1/T$. 
 At $\omega=1$ one has $q_0=1,\forall\beta$. In the zero-temperature limit 
 one has $q_0\to 1,\forall\omega$. The straight line is the infinite 
 temperature result. 
}
\label{saddle_clean}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
